---
title: "2. Epigenetic Analysis (CpG & GpC) Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Epigenetic Analysis (CpG & GpC) Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Welcome to the `sciNOME` Epigenetic Analysis pipeline. This module is designed to process both DNA Methylation (CpG) and Chromatin Accessibility (GpC) data from single-cell or bulk bisulfite sequencing (e.g., Bismark outputs). 

The workflow seamlessly guides you through genomic region aggregation, strict Quality Control (QC), missing value imputation, dimensionality reduction, and robust differential epigenetic analysis.

```{r setup, eval=FALSE}
# Load the package
library(sciNOME)
```

## 1. Data Aggregation from Bismark Files

The first step is to summarize base-resolution Bismark `.cov` (or `.cov.gz`) files into defined genomic regions (e.g., promoters, enhancers, or gene bodies) provided in a standard BED file.

The `Aggregate_epiRegions()` function reads all files in a directory and aggregates the methylated counts, unmethylated counts, and methylation levels in parallel.

```{r aggregate, eval=FALSE}
# Define paths
cov_directory <- "path/to/bismark_cov_files/"
bed_file_path <- "path/to/genomic_regions.bed"

# Run aggregation using 4 cores
raw_epi_mat <- Aggregate_epiRegions(
  cov_dir = cov_directory,
  bed_file = bed_file_path,
  n_cores = 4 
)
```

## 2. Quality Control (QC) and Matrix Extraction

Epigenetic matrices often contain missing values (NAs) due to low sequencing coverage. `sciNOME` provides a rigorous two-step QC method via `QC_epiData()`:
1. Retains the top *N* genomic regions with the lowest NA rates.
2. Removes samples with excessively high NA rates across these reliable regions.

```{r qc, eval=FALSE}
# Perform strict Two-Step QC
qc_epi_mat <- QC_epiData(
  data = raw_epi_mat,
  top_n_rows = 5000,           # Keep the top 5000 highest-quality regions
  max_col_na_ratio = 0.8       # Drop samples with >80% missing data
)
```

Once QC is complete, you might want to extract only the methylation `.level` matrix for clustering or visualization.

```{r extract, eval=FALSE}
# Extract ONLY the methylation level matrix (values between 0 and 1)
level_mat <- Extract_epiData(data = qc_epi_mat, suffix = ".level")
```

## 3. Dimensionality Reduction

Visualizing epigenetic data requires handling missing values. `Reduce_epiDims()` performs smart fuzzy matching with your metadata, imputes NAs (via KNN or Mean), and runs advanced dimensionality reduction algorithms (PCA, UMAP, MDS, or NMF).

```{r dimred, eval=FALSE}
# Assuming you have a metadata data.frame mapping sample IDs to conditions
# meta_df <- read.csv("metadata.csv")

epi_umap <- Reduce_epiDims(
  mat = level_mat,
  meta = meta_df,
  sample_col = "SampleID",
  group_col = "Condition",
  dr_method = "UMAP",          # Options: "PCA", "UMAP", "MDS", "NMF"
  impute_method = "knn",       # Advanced KNN imputation for missing values
  knn_k = 10
)

# The result is a data.frame ready for ggplot2 visualization
# head(epi_umap)
```

## 4. Differential Epigenetic Analysis

To find Differentially Methylated/Accessible Regions (DMRs/DARs), `Run_Diffanalysis()` uses an advanced statistical framework. It calculates variance and fold-change based on `.level` values, and performs **Fisher's Exact Test** using actual `.meth` and `.nonmeth` read counts for precise significance testing.

```{r diff_analysis, eval=FALSE}
# Example: Compare the "Disease" group against all other samples ("Rest")
diff_results <- Run_Diffanalysis(
  raw_mat = qc_epi_mat,          # Use the combined QC'd matrix (contains level, meth, nonmeth)
  meta = meta_df,                # Global metadata
  group_col = "Condition",       # Column defining the groups
  target_group = "Disease",      # Experimental group
  control_group = NULL,          # Set to NULL for "Target vs Rest", or specify a control like "Healthy"
  
  # Specify how your metadata maps to the matrix columns
  col_level = "Level_Col",       
  col_meth = "Meth_Col",
  col_nonmeth = "Unmeth_Col",
  
  effect_metric = "Diff",        # Metric to define direction (Options: "Diff", "logFC", "Scores")
  effect_th = 0.2,               # Minimum Absolute Difference threshold (e.g., 20% change)
  p_th = 0.05                    # P-value threshold
)

# The output table automatically classifies regions into "Up-regulated" or "Down-regulated"
# head(diff_results)
```

## Next Steps

Now that you have successfully processed both RNA and Epigenetic layers independently, you are ready to explore the cross-talk between them. Proceed to the next vignette: **3. Multi-Omics Integration**.
