---
title: "3. Multi-Omics Integration Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Multi-Omics Integration Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Welcome to the pinnacle of the `sciNOME` workflow: **Multi-Omics Integration**. 

Integrating Single-Cell RNA (scRNA-seq), DNA Methylation (CpG), and Chromatin Accessibility (GpC) is inherently difficult due to mismatched sample names and different feature types (Genes vs. Genomic Regions). The `Integrate_MultiOmics()` function solves this by using a **Global Metadata Table** to safely map samples, and a **Region Annotation Table** to bridge genomic coordinates with Gene IDs.

```{r setup, eval=FALSE}
# Load the package
library(sciNOME)
```

## 1. Preparing the Bridge: Region Annotation

To link epigenetic regions (e.g., `chr1:1000-2000`) to RNA expression (e.g., `ENSG000001`), we need a region annotation data.frame. It must contain the genomic coordinates and the associated gene.

```{r prep_region, eval=FALSE}
# Example of what the region_df should look like:
# chr    start    end      gene_id           gene_name
# chr1   14500    15500    ENSG00000223972   WASH7P
# chr1   68500    70500    ENSG00000237613   FAM138A

region_annotation <- read.csv("promoter_annotations.csv")
```

## 2. Basic TRI-Omics Integration

If you want to merge all three omics layers (RNA, CpG, GpC) for a specific biological group to see the average multi-omics landscape per gene, use `mode = "tri"`.

The algorithm automatically handles sample name mapping via `meta_df` and aggregates epigenetic region levels into gene-level averages.

```{r basic_tri, eval=FALSE}
# Merge all three layers for the "Disease" group
multi_omics_df <- Integrate_MultiOmics(
  mode = "tri",                 # Integrate all 3 omics
  target_group = "Disease",     # The biological condition to extract
  meta_df = global_meta,        # The master metadata linking all omics
  group_col = "Condition",      # Column in global_meta containing 'Disease'
  region_df = region_annotation,# The bridge table
  
  # 1. RNA info
  rna_obj = rna_obj,            # The processed RNA object from Vignette 1
  rna_id_col = "RNA_SampleID",  # Column in global_meta matching RNA cells
  
  # 2. CpG info
  cpg_mat = cpg_level_mat,      # Matrix from Vignette 2
  cpg_id_col = "CpG_SampleID",  # Column in global_meta matching CpG columns
  
  # 3. GpC info
  gpc_mat = gpc_level_mat,      # Matrix from Vignette 2
  gpc_id_col = "GpC_SampleID"   # Column in global_meta matching GpC columns
)

# Output contains GeneName, RNA_Exp, CpG_level, and GpC_level!
head(multi_omics_df)
```

## 3. Dual-Omics Integration

If you only have two omics layers, `sciNOME` seamlessly adapts using specific modes: `"rna_cpg"`, `"rna_gpc"`, or `"cpg_gpc"`.

```{r dual_omics, eval=FALSE}
# Example: Integrating only Epigenetic layers (Methylation & Accessibility)
epi_integration <- Integrate_MultiOmics(
  mode = "cpg_gpc",
  target_group = "Healthy",
  meta_df = global_meta,
  group_col = "Condition",
  region_df = region_annotation,
  
  cpg_mat = cpg_level_mat, cpg_id_col = "CpG_SampleID",
  gpc_mat = gpc_level_mat, gpc_id_col = "GpC_SampleID"
)
```

## 4. Advanced: Integration Driven by Differential Features

Often, you don't want to integrate the entire genome, but **only the features that are biologically significant** (e.g., DEGs, DMRs, DARs). 

`Integrate_MultiOmics` has a powerful built-in feature: you can pass your differential analysis results directly into the function. It will automatically apply your thresholds and subset the matrices *before* integration.

```{r advanced_diff, eval=FALSE}
# Assuming you have differential results from RunDEA_RNA() and Run_Diffanalysis()
integrated_sig_df <- Integrate_MultiOmics(
  mode = "tri",
  target_group = "Disease",
  meta_df = global_meta,
  group_col = "Condition",
  region_df = region_annotation,
  
  # RNA parameters with filtering
  rna_obj = rna_obj, rna_id_col = "RNA_SampleID",
  rna_diff_df = rna_deg_results,     # Pass RNA differential table
  rna_pval_col = "p_val", rna_pval_th = 0.05,
  rna_logfc_col = "avg_log2FC", rna_logfc_th = 0.25,
  
  # CpG parameters with filtering
  cpg_mat = cpg_level_mat, cpg_id_col = "CpG_SampleID",
  cpg_diff_df = cpg_dmr_results,     # Pass CpG differential table
  cpg_pval_col = "P.Value", cpg_pval_th = 0.05,
  cpg_diff_col = "Diff", cpg_diff_th = 0.15,
  
  # GpC parameters with filtering
  gpc_mat = gpc_level_mat, gpc_id_col = "GpC_SampleID",
  gpc_diff_df = gpc_dar_results,     # Pass GpC differential table
  gpc_pval_col = "P.Value", gpc_pval_th = 0.05,
  gpc_diff_col = "Diff", gpc_diff_th = 0.15
)

# The result is a highly specific multi-omics dataset of ONLY the significant cross-talks.
```

## Conclusion

With `sciNOME`, you have now successfully built a comprehensive data structure, rigorously processed RNA and Epigenetic matrices, and unified them into a multi-omics landscape. You can now use the `integrated_sig_df` output for advanced downstream plotting (e.g., scatter plots correlating expression vs. methylation).
