---
title: "1. Single-Cell RNA Analysis Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Single-Cell RNA Analysis Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Welcome to the `sciNOME` single-cell RNA sequencing analysis pipeline. This vignette demonstrates a complete, step-by-step workflow from raw count matrices to advanced downstream analysis, including dimensionality reduction, clustering, differential expression, and trajectory inference.

```{r setup, eval=FALSE}
# Load the package
library(sciNOME)
```

## 1. Data Preparation and Object Building

The first step is to initialize the `RNA` object. The `sciNOME` package provides a powerful one-step wrapper function, `Build_RNAObject()`, which automatically handles matrix formatting, NA replacement, pre-filtering, and metadata integration.

```{r build_object, eval=FALSE}
# Assuming you have a raw counts matrix (genes x cells) and sample metadata
# raw_counts <- read.csv("matrix.csv", row.names = 1)
# meta_df <- read.csv("metadata.csv", row.names = 1)

rna_obj <- Build_RNAObject(
  expr_mat = raw_counts, 
  meta_data = meta_df,
  min_cells = 3,           # Keep genes expressed in at least 3 cells
  min_features = 200,      # Keep cells with at least 200 detected genes
  mt_pattern = "^MT-",     # Regex pattern for mitochondrial genes (e.g., human)
  project_name = "Demo_RNA"
)
```

## 2. Quality Control (QC) and Normalization

Next, we filter out low-quality cells based on UMI counts, feature counts, and mitochondrial proportion. After filtering, `ProcessQC_RNA()` automatically performs sparse matrix normalization and scaling.

```{r qc_norm, eval=FALSE}
rna_obj <- ProcessQC_RNA(
  obj = rna_obj,
  mt_pattern = "^MT-",
  min_nCount = 500,
  max_nCount = 50000,
  min_nFeature = 200,
  max_nFeature = 8000,
  max_mt = 15,                  # Filter out cells with >15% mitochondrial reads
  norm_method = "LogNormalize", # Options: "LogNormalize", "LogCPM", or "TPM"
  do_scale = TRUE               # Center and scale the data for PCA
)
```
*Note: The filtered metadata and counts are stored in `filter_meta.data` and `filter_counts` safely without altering the raw input.*

## 3. Dimensionality Reduction

We then identify Highly Variable Genes (HVGs) and run Principal Component Analysis (PCA). The PCA coordinates are subsequently used to compute non-linear embeddings like t-SNE or UMAP.

```{r dim_red, eval=FALSE}
# 1. Run PCA (automatically calculates HVGs first)
rna_obj <- RunDimReduction_RNA(
  obj = rna_obj, 
  method = "PCA", 
  n_hvg = 2000, 
  pca_rank = 50
)

# 2. Run UMAP using the first 20 Principal Components
rna_obj <- RunDimReduction_RNA(
  obj = rna_obj, 
  method = "UMAP", 
  use_dims = 1:20
)
```

## 4. Cell Clustering

With the embeddings calculated, we group cells into clusters. `sciNOME` supports Graph-based (KNN + Leiden) and Hierarchical clustering approaches.

```{r clustering, eval=FALSE}
rna_obj <- RunClustering_RNA(
  obj = rna_obj,
  reduction = "pca",       # Use PCA space for distance calculation
  method = "graph",        # Graph-based Leiden clustering
  use_dims = 1:20,         
  cluster_res = 0.8,       # Resolution (higher = more clusters)
  k_neighbors = 20
)

# The cluster labels are automatically saved in:
# head(rna_obj$filter_meta.data$Auto_Cluster)
```

## 5. Differential Expression Analysis (DEA)

To identify marker genes for specific clusters, we perform an ultra-fast Wilcoxon Rank Sum Test using `RunDEA_RNA()`.

```{r dea, eval=FALSE}
# Example: Find marker genes for "Cluster_1" compared to all other cells
markers_cluster1 <- RunDEA_RNA(
  obj = rna_obj,
  layer_name = "data",
  group_col = "Auto_Cluster",
  ident_1 = "Cluster_1", 
  ident_2 = NULL,          # NULL implies "Rest" (all other groups)
  min_pct = 0.1,           # Minimum expression frequency
  logfc_thresh = 0.25      # Minimum log2 Fold Change
)

# View top significant genes
head(markers_cluster1)
```

## 6. Trajectory Inference (Pseudotime)

Finally, to explore cellular development or differentiation pathways, `sciNOME` offers pseudotime estimation. You must specify the starting (root) cluster.

```{r pseudotime, eval=FALSE}
rna_obj <- RunPseudotime_RNA(
  obj = rna_obj,
  reduction = "umap",
  group_col = "Auto_Cluster",
  start_clus = "Cluster_1",   # Define the biological root of the trajectory
  algorithm = "cluster"       # Options: "cluster" (Princurve) or "graph" (KNN shortest path)
)

# The pseudotime values are now merged into metadata:
# head(rna_obj$filter_meta.data$Pseudotime)
```

## Next Steps

Congratulations! You have successfully processed a single-cell RNA dataset. Proceed to the next vignettes to learn how to process Epigenetic data and integrate multiple omics layers!
